Meaningful data sharing

### 2020-09-18 new-paper.Rmd
notes from meeting

@@

Use a README file with your data project. A README is often the first place users will look to learn more about a directory [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards.

@@

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use a license from Creative Commons (<https://creativecommons.org/>), specifically, CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. Do not use non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

CC BY requires appropriate credit and attribution, but data can be shared and adapted, even for commercial use, provided changes are described. The CC0 is a "public domain" license, meaning data owners waive all rights, and the work is now "owned" by the public. CC0 data can be freely shared, copied, modified, and distributed, even for commercial purposes **without asking permission**. When using data with CC0, it is good practice to cite the original work, but is not required. Both CCBY and CC0 provide no warranty. Brief and full overviews of CCBY and CC0 can be found at [@ccby-short; @ccby-long; @cc0-short; @cc0-long].

Some data repositories and journals will automatically require a license when submitting data. For example, Dryad requires CC0 (<https://blog.datadryad.org/2011/10/05/why-does-dryad-use-cc0/> and <https://datadryad.org/stash/faq>) and the journal PLOS Comp Bio, requires licences for data that are not more restrictive than CC BY [@plos-comp-bio-data].

@@

**Variable names** are short, descriptive names with no spaces or special characters, such as "job_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details, see [@Broman2017; @Ellis2017; @Arslan2019].

Metadata should provided in plain text files following a standard format. Such formats include the Ecological Metadata Language [@EML-about] (EML), which uses XML (The eXtensible Markup Language), and Table Schema, which uses JSON-LD (Java Script Object Notation - Linked Data) \@ref(fig:json-ld). One implementation of this is Frictionless Data [@Fowler2017], which provides a language agnostic way to document data - to learn more see <https://frictionlessdata.io/guide/>.

EML is widely used in Ecology (for example [@Maerz2014]). JSON-LD provides advantages over EML, by having fields to facilitate data discovery with google dataset search [@google-data-search]. Use JSON-LD over EML, as it is more human readable, shorter, and quicker and easier to read and write. For more information on tooling to assist generating metadata, see <https://knb.ecoinformatics.org/tools/morpho> and [@Arslan2019].

```{r json-ld, fig.cap = "Example snippet of some Table Schema data for a dataset with three variables. This provides a description of each field, and the type of field."}
knitr::include_graphics("figures/carbon-json-ld.png")
```

@@

@@

As discussed in \@ref(cite),archive data using a platform generating an accession number such as a DOI. For sharing data collected for research, the most relevant homes are the research data repositories **zenodo**, **dryad**, or **Open Science Framework** (OSF). In addition to minting DOIs, they also provide citation templates that can be used in \@ref(cite).

Zenodo can be contacted to arrange special cases for storage of large data. Special hosting can be arranged via specially contacting Zenodo or your research institute. If using a research institute's library service to share data, ensure a DOI is provided so it can be accurately identified and cited.

There are many alternatives to these three, for example, an astronomy project may host its data at the Sloan Digital Sky Survey (SDSS), and genetic data can be hosted at GenBank [@GenBank]. Importantly, these all provide some form of accession number. Adhere to community standards when deciding where to publish data, as this will likely improve data reuse. Large data may also be shared via domain specific venues (e.g., in genomics or astronomy), where appropriate.

@@

@@

>>>>

**note: other little bit to say**

- "Here are some high level ideas on how to share data, that might obvious but are not often considered"

@@

Use a README file with your data project to communicate its purpose and contents. A README is often the first place users will look to learn more about a project [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards. For example, the data from Dryad, https://datadryad.org/stash/dataset/doi:10.5061/dryad.rn6k1 could be greatly enhanced by providing a README file that explains what the variables in the data collected are, and what the different sheets of the excel worksheet contain.

@@

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use the CC0 license, as it is commonly used, and provide simple rules for attribution, warranty, and use. The CC0 is a "public domain" license, meaning data owners waive all rights, and the work is now "owned" by the public. CC0 data can be freely shared, copied, modified, and distributed, even for commercial purposes **without asking permission**. When using data with CC0, it is good practice to cite the original work, but is not required. For more information on the benefit of using CC0, and why CCBY is not always appropriate, see https://osc.universityofcalifornia.edu/2016/09/cc-by-and-data-not-always-a-good-fit/.
Short and full overviews of CCBY and CC0 can be found at [@cc0-short; @cc0-long].

Do not use non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory (Figure \@ref(fig:dir-layout)). The license type should be mentioned in the README, if used.

@@

**Variable names** are short, descriptive names with no spaces or special characters, such as "job_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details, see [@Broman2017; @Ellis2017; @Arslan2019]. Metadata can be encoded into plain text formats that follow a set standard, such as EML or SCHEMA-LD. A deep understanding of these formats isn't required to use them, for more details, and guidance on generating them see @dataspice and [@Arslan2019].

@@

(**note: this bit needs some work**)
A trusted repository (such as Dryad) will manage citation for you. However, a CITATION file allows you to mention specifically how you want your data to be cited - this could point to the paper that used or described the data and collection methods, for example.

@@

(**note: on what to say**)
1. Data should go into Domain area - e.g., gbif for species data, or genbif or genbank for genetics data @GenBank. More information can be found at https://www.nature.com/sdata/policies/repositories. Importantly, domain areas should provide some form of accession number.

As discussed in \@ref(cite),archive data using a platform generating an accession number such as a DOI. For sharing data collected for research, the most relevant homes for general data are the research data repositories **zenodo** and **dryad**. In addition to minting DOIs, they also provide citation templates that can be used in \@ref(cite).

If using a research institute's library service to share data, ensure a DOI is provided so it can be accurately identified and cited.

@@

**mention bit about**

- https://www.the-scientist.com/news-opinion/top-spider-biologists-research-under-fire-67083
- https://www.nature.com/articles/d41586-020-00287-y

@@

**bit removed from this section**
